Retrofitting Effect Handlers onto OCaml

### 2021-03-14 retro-concurrency.tex
minor intro code edit, make contributions explicit

\vspace{-1em}
\subsection{Asynchronous I/O}
\label{sec:aio}

@@

\begin{lstlisting}[numbers=left]
let run main =
  let runq = Queue.create () in
  let suspend k =
    Queue.push (fun () -> continue k ()) runq
  in
  let pending_reads = ref [] in
  let rec run_next () =
    match Queue.pop runq with
    | f -> f ()
    | exception Queue.Empty ->
      match !pending_reads with
      | [] -> () (* no pending reads *)
      | todo ->
        let compl,todo = do_reads todo in
        List.iter (fun (str,k) ->
          Queue.push (fun () -> continue k str) runq) compl;
        pending_reads := todo;
        run_next ()
  in
  let rec spawn f =
    match f () with
    | () -> run_next () (* value case *)
    | effect Yield k -> suspend k; run_next ()
    | effect (Fork f') k -> suspend k; spawn f'
    | effect (Read ic) k ->
        pending_reads := (ic,k) :: !pending_reads;
        run_next ()
  in
  spawn main
\end{lstlisting}

The scheduler queue |runq| now holds thunks instead of continuations. The value
|pending_reads| maintains a list of pending reads and the associated
continuations (line 6). We handle the |Read| effect by pushing the pair of
input channel |ic| and continuation |k| to |pending_reads| (line 25), allowing
other threads in the scheduler to run.

When the scheduler queue is empty, the |run_next| function performs the pending
reads. We abstract away the details of the event-based I/O using the |do_reads|
function (line 14). |do_reads| takes a list of pending reads and blocks until
at least one of the reads succeeds. It returns a pair of lists |compl| and
|todo|. |compl| contains the result strings from successful reads and
corresponding continuations. These continuations are arranged to be resumed
with the read result and pushed into the scheduler queue. |todo| contains the
channels on which input is still pending and their corresponding continuations.
|pending_reads| is updated to point to the |todo| list so that they may be
attempted later. Observe that all of the changes to add asynchrony are
localised to the |run| function, and the computation that performs these
effects can remain in \emph{direct-style} (as opposed to the monadic-style in
Lwt and Async).

@@

\begin{itemize}
  \item the design and implementation of one-shot effect handlers for OCaml.
    Our design retains OCaml's compatibility with program analysis tools that
    inspect the stack using DWARF unwind tables. We have validated our DWARF
    unwind tables with the assistance of an automated validator
    tool~\cite{Bastian19}.
  \item a formal operational semantics for the effect handler implementation in
    OCaml.
	\item extensive evaluation which shows that our implementation has negligible
		impact on code that does not use effect handlers, and serves as an
		efficient foundation for scalable concurrent programming.
\end{itemize}

@@

\begin{figure*}
	\includegraphics[scale=0.35]{sandmark-notebook/sandmark_time}
	\caption{Normalized time of macro benchmarks. Baseline is Stock OCaml,
		whose running time in seconds in given in parenthesis.}
	\label{res:macro_time}
\end{figure*}

\begin{figure*}
	\includegraphics[scale=0.35]{sandmark-notebook/sandmark_codesize}
	\caption{Normalized OCaml text section size (OTSS) of macro benchmarks.
	Baseline is Stock OCaml, whose size in kilobytes given in parenthesis.}
	\label{res:macro_size}
\end{figure*}

@@

\begin{table}
\caption{Micro benchmarks without effects. Each entry is the percentage
	difference for Multicore OCaml over stock OCaml.}
%\vspace{-7mm}
{
\begin{tabular}{r c c c c c c c c c c}
	& \rot{exnval} & \rot{exnraise} & \rot{extcall} & \rot{callback} & \rot{ack}
	& \rot{fib} & \rot{motzkin} & \rot{sudan} & \rot{tak} \\ \hline
	\textbf{Time} & +0.0 & -1.9 & +17 & +65  & +5.3
								& +2.2 & +10 & +0.0 & +4.2 \\
	\textbf{Instr} & +0.0 & +0.0 & +10 & +72 & +16
								 & +24 & +16 & +14 & +17 \\ \hline
\end{tabular}
%\vspace{-5mm}
}
\label{tab:micro_noeffect}
\end{table}

@@

\begin{table}
\caption{Micro benchmarks with handlers but no perform. Each entry is the
	slowdown factor ($\times$ times) over its idiomatic implementation in stock OCaml.}
%\vspace{-3mm}
{
\begin{tabular}{r c c c c c}
	& \textbf{ack} & \textbf{fib} & \textbf{motzkin} & \textbf{sudan} & \textbf{tak} \\ \hline
	\texttt{MC} 	 	& 12.25 & 12.05 & 11.44 & 6.74 & 8.9 \\
	\texttt{monad} 	& 348.69 & 69.77 & 39.24 & 33.29 & 42.79 \\ \hline
\end{tabular}
}
\label{tab:micro_noperform}
	%\vspace{-5mm}
\end{table}

@@

\begin{figure}
	\begin{minipage}{0.49\linewidth}
		\centering
		\includegraphics[width=\linewidth]{http-benchmarks/throughput.pdf}
		%\vspace{-2mm}
		\subcaption{Throughput}
		\label{grf:throughput}
	\end{minipage}
	\begin{minipage}{0.49\linewidth}
		\includegraphics[width=\linewidth]{http-benchmarks/latency.pdf}
		%\vspace{-2mm}
		\subcaption{Tail latency}
		\label{grf:latency}
	\end{minipage}
	%\vspace{-3mm}
	\caption{Web server performance.}
	%\vspace{-5mm}
\end{figure}

>>>>

%\vspace{-1em}
\subsection{Asynchronous I/O}
\label{sec:aio}

@@

\begin{lstlisting}[numbers=left]
let run main =
  let runq = Queue.create () in
  let suspend k = Queue.push (continue k) runq in
  let pending_reads = ref [] in
  let rec run_next () =
    match Queue.pop runq with
    | f -> f ()
    | exception Queue.Empty ->
      match !pending_reads with
      | [] -> () (* no pending reads *)
      | todo ->
        let compl,todo = do_reads todo in
        List.iter (fun (str,k) ->
          Queue.push (fun () -> continue k str) runq) compl;
        pending_reads := todo;
        run_next ()
  in
  let rec spawn f =
    match f () with
    | () -> run_next () (* value case *)
    | effect Yield k -> suspend k; run_next ()
    | effect (Fork f') k -> suspend k; spawn f'
    | effect (Read ic) k ->
        pending_reads := (ic,k) :: !pending_reads;
        run_next ()
  in
  spawn main
\end{lstlisting}

The scheduler queue |runq| now holds thunks instead of continuations. The value
|pending_reads| maintains a list of pending reads and the associated
continuations (line 4). We handle the |Read| effect by pushing the pair of
input channel |ic| and continuation |k| to |pending_reads| (line 24), allowing
other threads in the scheduler to run.

When the scheduler queue is empty, the |run_next| function performs the pending
reads. We abstract away the details of the event-based I/O using the |do_reads|
function (line 12). |do_reads| takes a list of pending reads and blocks until
at least one of the reads succeeds. It returns a pair of lists |compl| and
|todo|. |compl| contains the result strings from successful reads and
corresponding continuations. These continuations are arranged to be resumed
with the read result and pushed into the scheduler queue. |todo| contains the
channels on which input is still pending and their corresponding continuations.
|pending_reads| is updated to point to the |todo| list so that they may be
attempted later. Observe that all of the changes to add asynchrony are
localised to the |run| function, and the computation that performs these
effects can remain in \emph{direct-style} (as opposed to the monadic-style in
Lwt and Async).

@@

\begin{itemize}
  \item the design and implementation of one-shot effect handlers for OCaml.
    Our design retains OCaml's compatibility with program analysis tools that
    inspect the stack using DWARF unwind tables. We have validated our DWARF
    unwind tables with the assistance of an automated validator
    tool~\cite{Bastian19}.
	\item a formal operational semantics for the effect handler implementation in
		OCaml. Our formalism explicitly models the subtle interactions with the C
		stack, which is generally overlooked by other formal models, but which the
		implementations must handle.
	\item extensive evaluation which shows that our implementation has negligible
		impact on code that does not use effect handlers, and serves as an
		efficient foundation for scalable concurrent programming.
\end{itemize}

@@

\begin{figure*}
	\includegraphics[scale=0.35]{sandmark-notebook/sandmark_time}
	\caption{Normalized time of macro benchmarks. Baseline is Stock OCaml,
		whose running time in seconds in given in parenthesis.}
	%\vspace{-2mm}
	\label{res:macro_time}
\end{figure*}

\begin{figure*}
	\includegraphics[scale=0.35]{sandmark-notebook/sandmark_codesize}
	\caption{Normalized OCaml text section size (OTSS) of macro benchmarks.
	Baseline is Stock OCaml, whose size in kilobytes given in parenthesis.}
	%\vspace{-2mm}
	\label{res:macro_size}
\end{figure*}

@@

\begin{table}
\caption{Micro benchmarks without effects. Each entry is the percentage
	difference for Multicore OCaml over stock OCaml.}
\vspace{-5mm}
{
\begin{tabular}{r c c c c c c c c c c}
	& \rot{exnval} & \rot{exnraise} & \rot{extcall} & \rot{callback} & \rot{ack}
	& \rot{fib} & \rot{motzkin} & \rot{sudan} & \rot{tak} \\ \hline
	\textbf{Time} & +0.0 & -1.9 & +17 & +65  & +5.3
								& +2.2 & +10 & +0.0 & +4.2 \\
	\textbf{Instr} & +0.0 & +0.0 & +10 & +72 & +16
								 & +24 & +16 & +14 & +17 \\ \hline
\end{tabular}
\vspace{-5mm}
}
\label{tab:micro_noeffect}
\end{table}

@@

\begin{table}
\caption{Micro benchmarks with handlers but no perform. Each entry is the
	slowdown factor ($\times$ times) over its idiomatic implementation in stock OCaml.}
\vspace{-3mm}
{
\begin{tabular}{r c c c c c}
	& \textbf{ack} & \textbf{fib} & \textbf{motzkin} & \textbf{sudan} & \textbf{tak} \\ \hline
	\texttt{MC} 	 	& 12.25 & 12.05 & 11.44 & 6.74 & 8.9 \\
	\texttt{monad} 	& 348.69 & 69.77 & 39.24 & 33.29 & 42.79 \\ \hline
\end{tabular}
}
\label{tab:micro_noperform}
\vspace{-3mm}
\end{table}

@@

\begin{figure}
	\begin{minipage}{0.49\linewidth}
		\centering
		\includegraphics[width=\linewidth]{http-benchmarks/throughput.pdf}
		%\vspace{-2mm}
		\subcaption{Throughput}
		\label{grf:throughput}
	\end{minipage}
	\begin{minipage}{0.49\linewidth}
		\includegraphics[width=\linewidth]{http-benchmarks/latency.pdf}
		%\vspace{-2mm}
		\subcaption{Tail latency}
		\label{grf:latency}
	\end{minipage}
	\vspace{-3mm}
	\caption{Web server performance.}
	\vspace{-5mm}
\end{figure}
